Combinatorial Proofs and Decomposition Theorems for First-order Logic\vspace{-3ex}

### 2021-01-26 main.tex
A few more fixes

The \bfit{complement} of a graph $\gG=\tuple{\vG,\eG}$ is
the graph $\compl\gG=\tuple{\vG,\compl\eG}$ where $vw\in\compl\eG$ iff
$vw\notin\eG$.

Let $\gG=\tuple{\vG,\eG}$ and $\gH=\tuple{\vH,\eH}$ be graphs such
that $\vG\cap\vH=\emptyset$. A \bfit{homomorphism}
$\phi\colon\gG\to\gH$ is a function $\phi\colon\vG\to\vH$ such that if
$vw\in\eG$ then $\phi(v)\phi(w)\in\eH$. The \bfit{union} $\gG +\gH$ is
the graph $\tuple{\vG \cup\vH,\eG\cup \eH}$ and the \bfit{join} $\gG
\times \gH$ is the graph $\tuple{\vG \cup\vH,\eG \cup \eH \cup \set{vw
\mid v \in \vG, w \in \vH}}$.  A graph $\gG$ is \bfit{disconnected}
if $\gG=\gG_1+\gG_2$ for two non-empty graphs $\gG_1,\gG_2$, otherwise
it is \bfit{connected}. It is \bfit{coconnected} if its complement is
connected.

@@

More formally, let $C$ be a formula with $\graphof C=\gC$, to which we
exhaustively apply the following
subformula rewriting steps, to obtain a sequent $\frameof C$:
\begin{enumerate}
\item {\bf Encode dependencies as fresh links.} For each dependency
  $\set{\single {x_i},\single {y_j}}$ in $\gC$, with corresponding
  subformulas $\exists x_i. A$ and $\forall y_j. B$ in $C$, we pick a
  fresh (nullary) predicate symbol $q_{i,j}$, and then replace $\exists
  x_i. A$ by $\dual q_{i,j} \cand \exists x_i. A$, and replace $\forall y_j. B$ by
  $q_{i,j} \cor \forall y_j. B$.
\item {\bf Erase quantifiers.} After step 1, remove all the
  quantifiers, i.e., replace $\exists x_i.A$ by $A$ and replace $\forall
  y_j.B$ by $B$ everywhere.
\item {\bf Simplify atoms.} After step 2, replace every predicate
  $p(t_1 \cdots t_n)$ (resp. $\dual p(t_1 \cdots t_n)$) with a nullary
  predicate symbol $p$ (resp.~$\dual p$)
%% \item {\bf Remove units.} All occurrences of $\ttt$ are replaced by
%%   $q_0\vlor\dual q_0$ and all $\fff$ are replaced by $q_0\vlan\dual
%%   q_0$ for some fresh nullary predicate $q_0$.
\end{enumerate}
Then $\linkingof{\frameof C}$ consists of the pairs induced by
$\linkingof\gC$ and the new pairs $\set{q_{i,j},\dual q_{i,j}}$ introduced in
step~1 above.
We call $\frameof C$ the \bfit{frame} of $C$ and we define the \bfit{frame} of $\gC$, denoted
$\frameof\gC$, as
$\tuple{\graphof{\frameof C},\linkingof{\frameof C}}$.

@@

\begin{remark}
  There is an alternative path of proving Theorem~\ref{thm:MLS1->fonet}
  by translating $\Deri$ to an $\FOMLL$-proof $\Pi$, observing that
  this process preserves the linking and the dualizer. However, for
  this, we have to extent the construction above to the $\cut$-rule,
  and then show that linking and dualizer of a sequent proof $\Pi$ are
  invariant under cut elimination. This can be done similarly to
  unification nets in~\cite{hughes:unifn}.
\end{remark}

@@

For each instance $\vlinf{\rr}{}{P}{Q}$ of an inference rule in
$\set{\wrD,\wfaD,\acD,\rectif\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we can
define the \bfit{induced map} $\mapof{\rr}\colon\vgraphof
Q\to\vgraphof P$ which acts as the identity for
$\rr\in\set{\me,\fequ}$ and as the canonical injection for
$\rr\in\set{\wrD,\wfaD}$. For $\rr=\acD$ it maps the vertices
corresponding to the two atoms in the premise to vertex of the
contracted atom in the conclusion, and for
$\rr\in\set{\rectif\cfaD,\rectif\mfaD,\rectif\mexD}$ it maps the two vertices
corresponding to the quantifiers in the premise to the one in the
conclusion (and acts as the identity on all other vertices).
For a
derivation $\Deri$ in
$\set{\wrD,\wfaD,\acD,\rectif\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we can then define
the \bfit{induced map} $\mapof\Deri$ as the composition of the induced
maps of the rule instances in $\Deri$.
%% \juihsuan{maybe mention at least that the induced
%% maps define graph homomorphisms. Do we need to talk about the contexts
%% $\Sconhole$ here (induced maps act clearly as the identity on contexts but we need them for the composition)?}
%% \lutz{For the context, I aleady say it is the identity. For the homom, it comes later}

@@

\begin{proof}[Proof of Lemma~\ref{lem:cw->skew}]
  First, observe that for every inference rule
  $\rr\in\set{\wrD,\wfaD,\acD,\rectif\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$
  the induced map $\mapof{\rr}\colon\vgraphof Q\to\vgraphof P$ defines
  a existential preserving graph homomorphism $\graphof Q\to\graphof
  P$ and a fibration on the corresponding binding graphs.  Therefore,
  their composition $\mapof\Deri$ has the same properties fibration.

@@

\begin{proof}
   By \cite[Proposition~7.6.1]{str:07:RTA}, we have a derivation
   $\vlder{\set{\wrD,\PE\fequ}}{\DDeri}{\PE H}{\PE G}$. In order to
   lift $\DDeri$, we might need to reorganize the instances of
   $\wD$. If $H$ contains a subformula $\forall x.A$ which is not
   present in $G$, the $\wD$-instances in $\DDeri$ could introduce the
   parts of the propositional encoding $x\vlor A$ independently.  We
   say that an instance $\rr_1$ of $\wD$ in $\Deri$ is \emph{in the
   scope} of an instance $\rr_2$ of $\wD$ if $\rr_1$ introduced
   formulas that contain a free variable $x$ (i.e., $x$ occurs in a
   term in a predicate) and $\rr_2$ introduces the atom $x$ as a
   subformula (i.e. the propositional encoding of the binder $x$).  We
   can now permute the $\wD$-instances in $\DDeri$ such that whenever
   a rule instance $\rr_1$ depends on an instance $\rr_2$, then
   $\rr_2$ occurs below $\rr_1$ in $\DDeri$.
   We can now lift $\DDeri$ stepwise. First, observe that each line of
   in $\DDeri$ is $\PE\fequ$-equivalent to the propositional encoding $\PE P$ of a
   first-order formula $P$.  We now have to show
   that each other instance of $\wrD$ in $\DDeri$ is indeed the image of a correct
   application of $\wD$ or $\wfaD$ in first-order logic.
   If we have a $\wD$ of the form
   \begin{equation*}
     \vlinf{\wD}{}{\PE S\cons{x\vlor \PE A}}{\PE S\cons{\PE A}}
     \qquor
     \vlinf{\wD}{}{\PE S\cons{(x\vlor \PE B)\vlor \PE A}}{\PE S\cons{\PE A}}
   \end{equation*}
   then $x$ cannot occur freely in $A$, as otherwise the fibration
   property would be violated. We can therefore lift these instances to
   \begin{equation*}
     \vlinf{\wfaD}{}{S\cons{\forall x. A}}{S\cons{A}}
     \qquor
     \vlinf{\wD}{}{S\cons{(\forall x. B)\vlor A}}{S\cons{A}}
   \end{equation*}
   respectively.     
  %% A subformula of $\PE B$ is called \emph{weak} if it has been
  %% introduced by a weakening. All subformulas of a weak formula are
  %% also weak. Two weak subformulas $B'$ and $B''$ of $\PE B$ form a
  %% \emph{weak pair} if $\PE B\fequ\Scons{B'\vlor B''}$ for some context
  %% $\Sconhole$. We can assume without loss of generality that
  %% whenever weak subformulas $B'$ and $B''$ form a weak pair, they have
  %% been introduced by the same instance of $\wrD$ in
  %% $\PE\Deri_1$.\footnote{If $\PE\Deri_1$ is not of that shape, it can
  %% brought into this form by simple rule permutations, and then
  %% collapsing several weakenings into a single one.}  Now we show that
  %% $\PE\Deri_1$ can be lifted. For this, observe that whenever a
  %% weakening in $\PE\Deri_1$ deletes an atom $x\in\VAR$, it must also
  %% delete all atoms in the scope of the corresponding quantifier,
  %% because $\phi'$ is a fibration on the binding graph. Hence, each line
  If a  weakening happens inside a subformula $x\vlor \PE C$ or $x\vlan
  \PE C$ in $\DDeri$, then there are the following cases:
  \begin{equation*}%\small
    \scalebox{.9}{$
    \vlinf{\wD}{}{\PE S\cons{x\vlor\PE  D\vlor \PE C}}{\PE S\cons{x\vlor\PE  C}}
    \quad
    \vlinf{\wD}{}{\PE S\cons{x\vlan (\PE D\vlor \PE C)}}{\PE S\cons{x\vlan\PE  C}}
    \quad
    \vlinf{\wD}{}{\PE S\cons{(x\vlor\PE  D)\vlan \PE C}}{\PE S\cons{x\vlan \PE C}}
    $}
  \end{equation*}
  The first two cases can be lifted to
  \begin{equation*}
    \vlinf{\wD}{}{S\cons{\forall x.( D\vlor C)}}{S\cons{\forall x.  C}}
    \quand
     \vlinf{\wD}{}{S\cons{\exists x.( D\vlor C)}}{S\cons{\exists x.  C}}
  \end{equation*}
  respectively. But in the third case,  an $\exists$-quantifier would be
  transformed into an $\forall$-quantifier. But as $\phi$ has to
  preserve existentials, this third case cannot occur. All other situations can be lifted trivially, giving us
  $\vlder{\set{\wrD,\wfaD,\fequ}}{\Deri}{H}{G}$ as desired.
\end{proof}

@@

The obvious next step in our research is to investigate proof
composition and normalisation of first-order combinatorial
proofs. Even in the propositional setting, the normalization of
combinatorial proofs is underdeveloped. There exist two different
procedures for cut elimination~\cite{hughes:invar,str:fscd17}, but
both have their insufficiencies, and there is no general theory.

Nonetheless, we hope to get new insights in the normalization of
classical first-order proofs through our work on combinatorial proofs.

@@

      instance of $\me$ can trivially be lifted, and and we can

>>>>

%The \bfit{complement} of a graph $\gG=\tuple{\vG,\eG}$ is
%the graph $\compl\gG=\tuple{\vG,\compl\eG}$ where $vw\in\compl\eG$ iff
%$vw\notin\eG$.

Let $\gG=\tuple{\vG,\eG}$ and $\gH=\tuple{\vH,\eH}$ be graphs such
that $\vG\cap\vH=\emptyset$. A \bfit{homomorphism}
$\phi\colon\gG\to\gH$ is a function $\phi\colon\vG\to\vH$ such that if
$vw\in\eG$ then $\phi(v)\phi(w)\in\eH$. The \bfit{union} $\gG +\gH$ is
the graph $\tuple{\vG \cup\vH,\eG\cup \eH}$ and the \bfit{join} $\gG
\times \gH$ is the graph $\tuple{\vG \cup\vH,\eG \cup \eH \cup \set{vw
\mid v \in \vG, w \in \vH}}$.  A graph $\gG$ is \bfit{disconnected}
if $\gG=\gG_1+\gG_2$ for two non-empty graphs $\gG_1,\gG_2$, otherwise
it is \bfit{connected}.
%It is \bfit{coconnected} if its complement is connected.

@@

More formally, let $C$ be a formula with $\graphof C=\gC$, to which we
exhaustively apply the following
subformula rewriting steps, to obtain a sequent $\frameof C$:
\begin{enumerate}
\item {\bf Encode dependencies as fresh links.} For each dependency
  $\set{\single {x_i},\single {y_j}}$ in $\gC$, with corresponding
  subformulas $\exists x_i. A$ and $\forall y_j. B$ in $C$, we pick a
  fresh (nullary) predicate symbol $q_{i,j}$, and then replace $\exists
  x_i. A$ by $\dual q_{i,j} \cand \exists x_i. A$, and replace $\forall y_j. B$ by
  $q_{i,j} \cor \forall y_j. B$.
\item {\bf Erase quantifiers.} After step 1, remove all the
  quantifiers, i.e., replace $\exists x_i.A$ by $A$ and replace $\forall
  y_j.B$ by $B$ everywhere.
\item {\bf Simplify atoms.} After step 2, replace every predicate
  $pt_1 \ldots t_n$ (resp. $\dual{p}t_1 \ldots t_n$) with a nullary
  predicate symbol $p$ (resp.~$\dual p$)
%% \item {\bf Remove units.} All occurrences of $\ttt$ are replaced by
%%   $q_0\vlor\dual q_0$ and all $\fff$ are replaced by $q_0\vlan\dual
%%   q_0$ for some fresh nullary predicate $q_0$.
\end{enumerate}
Then $\linkingof{\frameof C}$ consists of the pairs induced by
$\linkingof\gC$ and the new pairs $\set{q_{i,j},\dual q_{i,j}}$ introduced in
step~1 above.
We call $\frameof C$ the \bfit{frame} of $C$ and we define the \bfit{frame} of $\gC$, denoted
$\frameof\gC$, as
$\tuple{\graphof{\frameof C},\linkingof{\frameof C}}$.

@@

\begin{remark}
  There is an alternative path of proving Theorem~\ref{thm:MLS1->fonet}
  by translating $\Deri$ to an $\FOMLL$-proof $\Pi$, observing that
  this process preserves the linking and the dualizer. However, for
  this, we have to extend the construction above to the $\cut$-rule,
  and then show that linking and dualizer of a sequent proof $\Pi$ are
  invariant under cut elimination. This can be done similarly to
  unification nets in~\cite{hughes:unifn}.
\end{remark}

@@

For each instance $\vlinf{\rr}{}{P}{Q}$ of an inference rule in
$\set{\wrD,\wfaD,\acD,\rectif\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we can
define the \bfit{induced map} $\mapof{\rr}\colon\vgraphof
Q\to\vgraphof P$ which acts as the identity for
$\rr\in\set{\me,\fequ}$ and as the canonical injection for
$\rr\in\set{\wrD,\wfaD}$. For $\rr=\acD$ it maps the vertices
corresponding to the two atoms in the premise to the vertex of the
contracted atom in the conclusion, and for
$\rr\in\set{\rectif\cfaD,\rectif\mfaD,\rectif\mexD}$ it maps the two vertices
corresponding to the quantifiers in the premise to the one in the
conclusion (and acts as the identity on all other vertices).
For a
derivation $\Deri$ in
$\set{\wrD,\wfaD,\acD,\rectif\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we can then define
the \bfit{induced map} $\mapof\Deri$ as the composition of the induced
maps of the rule instances in $\Deri$.
%% \juihsuan{maybe mention at least that the induced
%% maps define graph homomorphisms. Do we need to talk about the contexts
%% $\Sconhole$ here (induced maps act clearly as the identity on contexts but we need them for the composition)?}
%% \lutz{For the context, I aleady say it is the identity. For the homom, it comes later}

@@

\begin{proof}[Proof of Lemma~\ref{lem:cw->skew}]
  First, observe that for every inference rule
  $\rr\in\set{\wrD,\wfaD,\acD,\rectif\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$
  the induced map $\mapof{\rr}\colon\vgraphof Q\to\vgraphof P$ defines
  an existential-preserving graph homomorphism $\graphof Q\to\graphof
  P$ and a fibration on the corresponding binding graphs.  Therefore,
  their composition $\mapof\Deri$ has the same properties of fibration.

@@

\begin{proof}
   By \cite[Proposition~7.6.1]{str:07:RTA}, we have a derivation
   $\vlder{\set{\wrD,\PE\fequ}}{\DDeri}{\PE H}{\PE G}$. In order to
   lift $\DDeri$, we might need to reorganize the instances of
   $\wD$. If $H$ contains a subformula $\forall x.A$ which is not
   present in $G$, the $\wD$-instances in $\DDeri$ could introduce the
   parts of the propositional encoding $x\vlor A$ independently.  We
   say that an instance $\rr_1$ of $\wD$ in $\Deri$ is \emph{in the
   scope} of an instance $\rr_2$ of $\wD$ if $\rr_1$ introduced
   formulas that contain a free variable $x$ (i.e., $x$ occurs in a
   term in a predicate) and $\rr_2$ introduces the atom $x$ as a
   subformula (i.e. the propositional encoding of the binder $x$).  We
   can now permute the $\wD$-instances in $\DDeri$ such that whenever
   a rule instance $\rr_1$ depends on an instance $\rr_2$, then
   $\rr_2$ occurs below $\rr_1$ in $\DDeri$.
   We can now lift $\DDeri$ stepwise. First, observe that each line of
   $\DDeri$ is $\PE\fequ$-equivalent to the propositional encoding $\PE P$ of a
   first-order formula $P$.  We now have to show
   that each instance of $\wrD$ in $\DDeri$ is indeed the image of a correct
   application of $\wD$ or $\wfaD$ in first-order logic.
   If we have a $\wD$ of the form
   \begin{equation*}
     \vlinf{\wD}{}{\PE S\cons{x\vlor \PE A}}{\PE S\cons{\PE A}}
     \qquor
     \vlinf{\wD}{}{\PE S\cons{(x\vlor \PE B)\vlor \PE A}}{\PE S\cons{\PE A}}
   \end{equation*}
   then $x$ cannot occur freely in $A$, as otherwise the fibration
   property would be violated. We can therefore lift these instances to
   \begin{equation*}
     \vlinf{\wfaD}{}{S\cons{\forall x. A}}{S\cons{A}}
     \qquor
     \vlinf{\wD}{}{S\cons{(\forall x. B)\vlor A}}{S\cons{A}}
   \end{equation*}
   respectively.     
  %% A subformula of $\PE B$ is called \emph{weak} if it has been
  %% introduced by a weakening. All subformulas of a weak formula are
  %% also weak. Two weak subformulas $B'$ and $B''$ of $\PE B$ form a
  %% \emph{weak pair} if $\PE B\fequ\Scons{B'\vlor B''}$ for some context
  %% $\Sconhole$. We can assume without loss of generality that
  %% whenever weak subformulas $B'$ and $B''$ form a weak pair, they have
  %% been introduced by the same instance of $\wrD$ in
  %% $\PE\Deri_1$.\footnote{If $\PE\Deri_1$ is not of that shape, it can
  %% brought into this form by simple rule permutations, and then
  %% collapsing several weakenings into a single one.}  Now we show that
  %% $\PE\Deri_1$ can be lifted. For this, observe that whenever a
  %% weakening in $\PE\Deri_1$ deletes an atom $x\in\VAR$, it must also
  %% delete all atoms in the scope of the corresponding quantifier,
  %% because $\phi'$ is a fibration on the binding graph. Hence, each line
  If a  weakening happens inside a subformula $x\vlor \PE C$ or $x\vlan
  \PE C$ in $\DDeri$, then there are the following cases:
  \begin{equation*}%\small
    \scalebox{.9}{$
    \vlinf{\wD}{}{\PE S\cons{x\vlor\PE  D\vlor \PE C}}{\PE S\cons{x\vlor\PE  C}}
    \quad
    \vlinf{\wD}{}{\PE S\cons{x\vlan (\PE D\vlor \PE C)}}{\PE S\cons{x\vlan\PE  C}}
    \quad
    \vlinf{\wD}{}{\PE S\cons{(x\vlor\PE  D)\vlan \PE C}}{\PE S\cons{x\vlan \PE C}}
    $}
  \end{equation*}
  The first two cases can be lifted to
  \begin{equation*}
    \vlinf{\wD}{}{S\cons{\forall x.( D\vlor C)}}{S\cons{\forall x.  C}}
    \quand
     \vlinf{\wD}{}{S\cons{\exists x.( D\vlor C)}}{S\cons{\exists x.  C}}
  \end{equation*}
  respectively. But in the third case,  an $\exists$-quantifier would be
  transformed into an $\forall$-quantifier. But as $\phi$ has to
  preserve existentials, this third case cannot occur. All other situations can be lifted trivially, giving us
  $\vlder{\set{\wrD,\wfaD,\fequ}}{\Deri}{H}{G}$ as desired.
\end{proof}

@@

The obvious next step in our research is to investigate proof
composition and normalisation of first-order combinatorial
proofs. Even in the propositional setting, the normalisation of
combinatorial proofs is underdeveloped. There exist two different
procedures for cut elimination~\cite{hughes:invar,str:fscd17}, but
both have their insufficiencies, and there is no general theory.

Nonetheless, we hope to get new insights in the normalisation of
classical first-order proofs through our work on combinatorial proofs.

@@

      instance of $\me$ can trivially be lifted, and we can
